---
title: "Homework Assignment #2"
subtitle: "Exploring patterns of environmental justice"
author: "Ryan Green"
date: 10/19/2024
execute: 
  eval: true
  warning: false
  message: false
  results: hide
format:
  html:
    toc: true
editor_options: 
  chunk_output_type: console
---

Loading libraries

```{r load_libraries}
library(sf) 
library(tmap)
library(here)
library(terra)
library(stars)
library(stringr)
library(testthat)
library(tidyverse)
library(RColorBrewer)
```

Loading in all data: EJScreen, LA Redlined Districts, LA Biodiversity

```{r load_data}
ejscreen <- sf::st_read(here::here("data", "ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"), quiet = TRUE) 

redlining <- sf::st_read(here::here("data", "mapping-inequality", "mapping-inequality-los-angeles.json"), quiet = TRUE)

biodiversity <- sf::st_read(here::here("data", "gbif-birds-LA", "gbif-birds-LA.shp"), quiet = TRUE)
```

## Part 1: Legacy of redlining in current environmental (in)justice

Isolating Los Angeles County and surrounding counties for basemap

```{r isolate_counties}
los_angeles <- ejscreen %>%
  dplyr::filter(CNTY_NAME %in% c("Los Angeles County"))
orange <- ejscreen %>%
  dplyr::filter(CNTY_NAME %in% c("Orange County"))
san_bernardino <- ejscreen %>%
  dplyr::filter(CNTY_NAME %in% c("San Bernardino County"))
```

Checking for invalid polygons

```{r invalid_polygon_test}
redlining_valid <- st_is_valid(redlining)
valid <- all(redlining_valid)

if (valid == TRUE) {
  print("No invalid polygons")
} else {
  print("Contains invalid polygons")
}
```

Transforming CRS for `redlining` and invalid polygons, and checking CRS with `los_angeles`

```{r crs_transform}
redlining <- st_transform(redlining, crs = 3857)

redlining_valid <- st_is_valid(redlining)
valid <- all(redlining_valid)

if (valid == TRUE) {
  print("No invalid polygons")
} else {
  print("Contains invalid polygons")
}
```

Testing the CRS transform

```{r crs_test}
message("Running CRS check...")
expect_identical(st_crs(los_angeles) == st_crs(redlining), TRUE, label = "true", expected.label = "false")
message("CRS are identical")
```

Calculating the percentages of HOLC district rankings

```{r holc_grade_percentage_table}
census_block <- st_join(los_angeles, redlining, join = st_intersects, left = TRUE) %>%
  group_by(grade) %>% 
  summarise(grade_percent = (n()/nrow(los_angeles))*100) %>%
  select(grade, grade_percent) %>%
  st_drop_geometry()

kableExtra::kable(census_block,
                  format = "simple",
                  col.names = c("HOLC Grade",
                                "Percentage"),
                  caption = "Grade Percentages of Redlined Districts",
                  align = "l",
                  digits = 1)
```

Creating a table for the current EJ conditions of redlined districts, noting PM2.5 levels, mean Low Income percentage, and Mean Life Expectancy percentage.

```{r ej_conditions_table}
ej_conditions <- st_join(los_angeles, redlining, join = st_intersects, left = TRUE) %>%
  group_by(grade) %>% 
  summarise(
    mean_pm25 = mean(PM25, na.rm = TRUE),
    mean_lowincpct = (mean(LOWINCPCT, na.rm = TRUE)*100),
    mean_lifeexppct = (mean(LIFEEXPPCT, na.rm = TRUE)*100)) %>%
  st_drop_geometry()

kableExtra::kable(ej_conditions, 
                  format = "simple", 
                  col.names = c("HOLC Grade", 
                                "Mean PM2.5 Levels", 
                                "Mean Low Income %", 
                                "Mean Life Expectancy %"), 
                  caption = "Current EJ Conditions Within HOLC Grades",
                  align = "l",
                  digits = 1)
```

Manipulating the redlined districts' EJ data and creating a bar plot

```{r ej_conditions_barplot}
ej_conditions_long <- ej_conditions %>%
  pivot_longer(cols = c("mean_pm25",
                        "mean_lowincpct",
                        "mean_lifeexppct"),
               names_to = "variable", 
               values_to = "mean_value")

ggplot(ej_conditions_long, aes(x = grade, 
                               y = mean_value, 
                               fill = variable)) +
  geom_bar(stat = "identity", 
           position = "dodge") +
  labs(x = "Grade", 
       y = "Mean Values", 
       fill = "Variable") +
  scale_fill_manual(values = c("mean_pm25" = "skyblue3", 
                               "mean_lowincpct" = "maroon", 
                               "mean_lifeexppct" = "springgreen3"),
                    labels = c("mean_pm25" = "Mean PM25", 
                               "mean_lowincpct" = "Mean Low Income %", 
                               "mean_lifeexppct" = "Mean Low Life Expectancy %")) +
  theme_light()
```

These results show that both low income and low life expectancy percentages increase with a decreasing HOLC grade. Low income levels are more than twice as high in grade D districts than in grade A. PM2.5 levels are consistent across all districts, which is to be expected. Overall, these results align with the HOLC ranking, unfortunately. 

----

Creating a line object based on `los_angeles`, in order to change line color
Creating Map of redlined districts in Los Angeles county

```{r la_redline_map}
la_borders <- los_angeles %>% sf::st_cast("MULTILINESTRING")
orange_borders <- orange %>% sf::st_cast("MULTILINESTRING")
sb_borders <- san_bernardino %>% sf::st_cast("MULTILINESTRING")

tm_shape(la_borders, 
         bbox = redlining) +
  tm_lines(lwd = 1.5, 
           col = "#665A47", 
           alpha = 0.4) +
tm_shape(redlining) +
  tm_polygons(title = "HOLC Grade",
              col = "grade",
              palette = c("springgreen4", 
                          "dodgerblue4", 
                          "goldenrod1", 
                          "firebrick3"),
              alpha = 0.8,
              legend.is.portrait = FALSE) +
tm_shape(orange_borders) +
  tm_lines(lwd = 1, 
           col = "#665A47", 
           alpha = 0.1) +
tm_shape(sb_borders) +
  tm_lines(lwd = 1, 
           col = "#665A47", 
           alpha = 0.1) +
tm_layout(bg.color = "#F0EBD8",
          title = "LOS ANGELES", 
          title.fontfamily = "times", 
          title.fontface = "bold", 
          title.size = 1.6,
          title.position = c(0.67, 0.23),
          title.color = "#493E2E",
          legend.position = c(0.64, 0),
          legend.frame = TRUE,
          legend.frame.lwd = 1.2,
          legend.bg.color = "#F6EAC7",
          legend.title.fontfamily = "times",
          legend.title.fontface = "bold",
          legend.text.color = "#493E2E",
          legend.title.color = "#493E2E",
          legend.bg.alpha = 0.5) +
  tm_credits("Historic Redlined Districts",
             size = 1,
             col = "#493E2E",
             fontfamily = "times",
             fontface = "bold",
             position = c(0.66, 0.16)) +
  tm_compass(type = "rose",
             position = c(0.03, 0.03),
             color.dark = "#665A47",
             color.light = "#F6EAC7",
             size = 3) +
  tm_scale_bar(text.color = "#665A47",
               color.dark = "#665A47",
               color.light = "#F6EAC7",
               position = c(0.67, 0.13))
```


## Part 2: Legacy of redlining in biodiversity observations

Creating joined dataframe of biodiversity and redlining datasets

```{r redline_bio_join}
biodiversity <- st_transform(biodiversity, crs = 3857)

bio_red_join <- st_join(biodiversity, redlining, join = st_intersects, left = TRUE)
```

Creating table of percentages of bird observations, by HOLC grade

```{r obs_per_grade_table}
bio_red <- bio_red_join %>%
  filter(year == 2022) %>%
  group_by(grade) %>%
  filter(grade != "NA") %>%
  summarise(count_id = n()) %>%
  mutate(bird_grade_percent = count_id/sum(count_id)*100) %>%
  select(grade, bird_grade_percent) %>%
  st_drop_geometry()

kableExtra::kable(bio_red,
                  format = "simple", 
                  caption = "Percentage of Bird Observations Within Redlined Districts, 2022",
                  col.names = c("Grade", "Percentage of Observations"),
                  align = "l",
                  digits = 1)

```

Plotting the results from the above table into a barplot

```{r obs_percentage_barplot}
ggplot(bio_red, aes(x = grade, y = bird_grade_percent)) +
         (geom_col(fill = "springgreen4")) +
  labs(title = "Percentage of Bird Observations by HOLC Grade in 2022",
       x = "HOLC Grade",
       y = "Percentage of Bird Observations") +
  theme_light()
```


The results show that in 2022, both grades C and D have a higher percentage of bird observations than grades A and B. These results seem to contradict the assigned ranking by the HOLC, as lower rankings are typically associated with fewer green areas and biodiversity. However, this is not a comprehensive look at the biodiversity of those districts, as we only use the data for 2022. 

Soto et al. 2023 found that districts with lower rankings have lower numbers of bird observations not because of lower inherent biodiversity, but by a disparity of data about observations in those districts. However our dataset shows otherwise, as grade C has by far the most observations of any grade (47,973 observations). While this is inconsistent with the findings of Soto et al., it is also important to note that our biodiversity dataset may be skewed for the same reason: greater numbers of observations does not necessarily mean greater biodiversity. 


----

----

----


Below, I created another figure and a map of bird observations, just for fun. This is not part of the homework but I thought I would include it. 


Creating another table of all the biodiversity data (2021-2023)

```{r all_bio_table}
bio_red_all <- bio_red_join %>%
  group_by(grade) %>%
  filter(grade != "NA") %>%
  summarise(count_id = n()) %>%
  mutate(bird_grade_percent = count_id/sum(count_id)*100) %>%
  select(grade, bird_grade_percent) %>%
  st_drop_geometry()

kableExtra::kable(bio_red_all,
                  format = "simple", 
                  caption = "Percentage of Bird Observations Within Redlined Districts, 2021-2023",
                  col.names = c("Grade", "Percentage of Observations"),
                  align = "l",
                  digits = 1)

ggplot(bio_red_all, aes(x = grade, y = bird_grade_percent)) +
         (geom_col(fill = "springgreen4")) +
  labs(title = "Percentage of Bird Observations by HOLC Grade, 2021-2023",
       x = "HOLC Grade",
       y = "Percentage of Bird Observations") +
  theme_light()

```

Counting the total number of observations in each grade from the entire biodiversity dataset

```{r total_obs_bio}
count_bio <- bio_red_join %>%
  group_by(grade) %>%
  summarise(count = n()) %>%
  st_drop_geometry()

kableExtra::kable(count_bio, 
                  format = "simple", 
                  col.names = c("Grade", "Total Observations"), 
                  caption = "Total Observations by Grade",
                  align = "l")

```

Adding the total bird observations from each district in 2022 into a column on the redlined districts dataset, for plotting

```{r bird_obs_column_redlined}
la_bio <- bio_red_join %>%
  filter(year == 2022) %>%
  filter(grade != "NA")

intersections <- st_intersects(redlining, la_bio)

redlining$bird_count <- sapply(intersections, length)
```

Creating map of bird observations in redlined districts in 2022

```{r bird_obs_map}
custom_breaks <- c(0, 51, 151, 251, 350)

tm_shape(redlining) +
  tm_polygons(title = " ",
              col = "bird_count",
              breaks = custom_breaks,
              palette = c("#E6E696", "#0E4209"),
              legend.show = FALSE) +
tm_shape(la_borders) +
  tm_lines(lwd = 1, 
           col = "#665A47", 
           alpha = 0.3) +
tm_shape(orange_borders) +
  tm_lines(lwd = 1, 
           col = "#665A47", 
           alpha = 0.1) +
tm_shape(sb_borders) +
  tm_lines(lwd = 1, 
           col = "#665A47", 
           alpha = 0.1) +
tm_layout(bg.color = "#F0EBD8",
          title = "BIRD OBSERVATIONS", 
          title.fontfamily = "times", 
          title.fontface = "bold", 
          title.size = 1.2,
          title.position = c(0.63, 0.20),
          title.color = "#493E2E") +
tm_add_legend(type = "fill", 
              labels = c("0-50", 
                         "51-150", 
                         "151-250", 
                         "251-350"),
                col = c("#E6E796", 
                        "#9EB067", 
                        "#557937", 
                        "#0E410B" ),
              border.col = "#493E2E",
                title = " ",
                is.portrait = FALSE,
              alpha = 0.7) +
  tm_legend(position = c(0.65, 0.015),
            text.color = "#493E2E") +
  tm_credits("In Historically Redlined Districts",
             size = 1,
             col = "#493E2E",
             fontfamily = "times",
             fontface = "italic",
             position = c(0.615, 0.13)) +
  tm_compass(type = "rose",
             position = c(0.03, 0.03),
             color.dark = "#665A47",
             color.light = "#F6EAC7",
             size = 3) +
  tm_scale_bar(text.color = "#665A47",
               color.dark = "#665A47",
               color.light = "#F6EAC7",
               position = c(0.65, 0.09))



```
